¥documentclass[11pt]{article}
¥usepackage{amsmath, amssymb, booktabs, hyperref}
¥usepackage[margin=1in]{geometry}

¥title{Mixed-Effects Analysis of Multi-Trial Forecasting Results¥¥
on ForecastBench Market Questions}
¥author{Kevin Murphy (with help from Claude Code)}
¥date{6 April 2026}

¥begin{document}
¥maketitle

¥section{Introduction}

We evaluate an LLM-based agentic forecaster on 263 binary market questions
from ForecastBench, drawn from four prediction market sources:
Polymarket (136), Manifold (68), Metaculus (50), and Infer (9).
Questions were asked between 2025-10-26 and 2026-03-31 (after the LLM
knowledge cutoff to prevent parametric leakage).

Each configuration is run $K=5$ independent trials per question, with
logit-space shrinkage aggregation across trials.
We compare three methods:
¥begin{enumerate}
  ¥item ¥textbf{Pro+search}: Gemini‾3.1‾Pro with high thinking, Brave web
    search, crowd estimate visible, source-specific tools enabled.
  ¥item ¥textbf{Flash+search}: Gemini‾3‾Flash with the same settings.
  ¥item ¥textbf{Pro+nosearch}: Gemini‾3.1‾Pro with high thinking, no web
    search, crowd estimate visible, no source-specific tools.
¥end{enumerate}

This design isolates two effects: (a)‾model quality (Pro vs.¥ Flash,
both with search) and (b)‾the value of web search (Pro with vs.¥ without).


¥section{Model}

We fit a two-way fixed-effects model to the per-question, per-method scores:
¥begin{equation}
  ¥mathrm{score}(q, m, k) = ¥mu + ¥alpha_m + ¥gamma_q + ¥varepsilon_{q,m,k},
  ¥label{eq:model}
¥end{equation}
where:
¥begin{itemize}
  ¥item $¥mu$ is the grand mean score across all methods and questions,
  ¥item $¥alpha_m$ is the ¥emph{method effect} for method $m$ (what we want to estimate),
  ¥item $¥gamma_q$ is the ¥emph{question difficulty} for question $q$ (nuisance parameter),
  ¥item $¥varepsilon_{q,m,k}$ is the residual trial noise.
¥end{itemize}

We use two views of the data:

¥paragraph{Aggregated scores (for method effects).}
For each (question‾$q$, method‾$m$) pair, we have one aggregated score:
¥begin{equation}
  s_{q,m} = ¥ell¥!¥bigl(¥hat{p}_{q,m},¥; o_q¥bigr),
¥end{equation}
where $¥hat{p}_{q,m}$ is the aggregated forecast, $o_q ¥in ¥{0,1¥}$ is the
outcome, and $¥ell$ is the scoring function (Metaculus or Brier).

We fit model‾¥eqref{eq:model} to the $Q ¥times M$ aggregated scores
via alternating projections (block coordinate descent):
¥begin{align}
  ¥gamma_q &¥leftarrow ¥frac{1}{M} ¥sum_m (s_{q,m} - ¥mu - ¥alpha_m), ¥¥
  ¥alpha_m &¥leftarrow ¥frac{1}{Q} ¥sum_q (s_{q,m} - ¥mu - ¥gamma_q),
¥end{align}
until convergence (typically $< 50$ iterations).

¥paragraph{Trial-level scores (for ANOVA).}
To separate within-trial noise from between-question difficulty,
we use all $K ¥times M ¥times Q$ individual trial scores:
¥begin{equation}
  s_{q,m,k} = ¥ell(p_{q,m,k},¥; o_q), ¥quad k = 1, ¥ldots, K,
¥end{equation}
where $p_{q,m,k}$ is the raw forecast from trial‾$k$ (before aggregation).
We perform a Type‾III sum-of-squares decomposition:
¥begin{align}
  ¥mathrm{SS}_{¥mathrm{total}} &= ¥sum_{q,m,k} (s_{q,m,k} - ¥bar{s})^2, ¥¥
  ¥mathrm{SS}_{¥mathrm{method}} &= KQ ¥sum_m (¥bar{s}_{¥cdot m ¥cdot} - ¥bar{s})^2, ¥¥
  ¥mathrm{SS}_{¥mathrm{question}} &= KM ¥sum_q (¥bar{s}_{q ¥cdot ¥cdot} - ¥bar{s})^2, ¥¥
  ¥mathrm{SS}_{¥mathrm{residual}} &= ¥mathrm{SS}_{¥mathrm{total}}
    - ¥mathrm{SS}_{¥mathrm{method}} - ¥mathrm{SS}_{¥mathrm{question}}.
¥end{align}


¥section{Results: Metaculus Score}

¥subsection{Method effects}

The grand mean across all three methods is $¥hat{¥mu} = 67.5$.

¥begin{table}[h]
¥centering
¥begin{tabular}{lrr}
¥toprule
Method & Effect ($¥hat{¥alpha}_m$) & Adjusted score ¥¥
¥midrule
Pro+search      & $+4.8$ & $72.3$ ¥¥
Flash+search    & $+1.4$ & $68.9$ ¥¥
Pro+nosearch    & $-6.2$ & $61.4$ ¥¥
¥bottomrule
¥end{tabular}
¥caption{Method effects for Metaculus score (higher = better).
  The adjusted score is $¥hat{¥mu} + ¥hat{¥alpha}_m$.}
¥label{tab:metaculus_effects}
¥end{table}

¥subsection{Pairwise comparisons}

We compare each method to the best (Pro+search) using paired differences
on the same 263 questions, with bootstrap $p$-values.
For each question $q$ we compute
$¥Delta_q = s_{q,m} - s_{q,¥mathrm{ref}}$, then bootstrap the mean
of $¥Delta_q$ over 2{,}000 resamples.  The quantity $p(¥Delta > 0)$ is the
fraction of bootstrap replicates where the mean difference is positive
(i.e., the method beats the reference).

¥begin{table}[h]
¥centering
¥begin{tabular}{lrrr}
¥toprule
Method vs.¥ Pro+search & $¥Delta$ mean & $¥Delta$ std & $p(¥Delta > 0)$ ¥¥
¥midrule
Pro+nosearch     & $-11.0$ & $42.8$ & $0.000$ ¥¥
Flash+search     & $-3.4$  & $30.0$ & $0.011$ ¥¥
¥bottomrule
¥end{tabular}
¥caption{Pairwise comparisons vs.¥ Pro+search (Metaculus score, $n=263$).
  $p(¥Delta > 0)$ is the bootstrap probability that the method beats
  the reference.}
¥label{tab:metaculus_pairwise}
¥end{table}

Both comparisons are statistically significant (Table‾¥ref{tab:metaculus_pairwise}).
The effect of web search is large ($¥Delta = -11.0$ without search,
$p < 0.001$), and Pro outperforms Flash by a smaller but significant margin
($¥Delta = -3.4$, $p = 0.011$).


¥section{Results: Brier Score}

¥subsection{Method effects}

The grand mean Brier score is $¥hat{¥mu} = 0.0625$.

¥begin{table}[h]
¥centering
¥begin{tabular}{lrr}
¥toprule
Method & Effect ($¥hat{¥alpha}_m$) & Adjusted score ¥¥
¥midrule
Pro+search      & $-0.0105$ & $0.0519$ ¥¥
Flash+search    & $-0.0032$ & $0.0593$ ¥¥
Pro+nosearch    & $+0.0137$ & $0.0761$ ¥¥
¥bottomrule
¥end{tabular}
¥caption{Brier score method effects (lower = better).}
¥label{tab:brier_effects}
¥end{table}

The ranking is identical to the Metaculus score analysis.

¥subsection{Variance decomposition}

¥begin{table}[h]
¥centering
¥begin{tabular}{lrrrr}
¥toprule
Source & SS & ¥% & df & $F$ ¥¥
¥midrule
Method   & $6.72$  & $5.0¥%$  & 2    & $822^*$ ¥¥
Question & $113.92$ & $84.0¥%$ & 262  & $106^*$ ¥¥
Residual & $15.03$ & $11.1¥%$ & 3{,}680 & --- ¥¥
¥midrule
Total    & $135.66$ & $100¥%$ & $3{,}944$ & ¥¥
¥bottomrule
¥end{tabular}
¥caption{ANOVA for 3 methods ($K=5$, $Q=263$).
  $^*$Significant at $p < 0.001$.
  $F_{¥mathrm{crit}}(2, 3680) ¥approx 3.0$.}
¥label{tab:brier_anova}
¥end{table}

¥textbf{Question difficulty explains $84¥%$ of total variance}
(Table‾¥ref{tab:brier_anova}), trial-to-trial stochasticity explains
$11¥%$, and method differences account for $5¥%$.  Unlike the AIBQ2
analysis (where method effects were not significant with 4 similar methods),
here the method effect is highly significant ($F = 822$, $p < 0.001$)
because the three methods span a much wider performance range
(with vs.¥ without search).

¥subsection{Pairwise comparisons}

¥begin{table}[h]
¥centering
¥begin{tabular}{lrrr}
¥toprule
Method vs.¥ Pro+search & $¥Delta$ mean & $¥Delta$ std & $p(¥Delta > 0)$ ¥¥
¥midrule
Pro+nosearch     & $+0.0242$ & $0.1019$ & $0.000$ ¥¥
Flash+search     & $+0.0074$ & $0.0696$ & $0.035$ ¥¥
¥bottomrule
¥end{tabular}
¥caption{Pairwise comparisons vs.¥ Pro+search (Brier score, $n=263$).
  Positive $¥Delta$ means the method is worse (higher Brier score).}
¥label{tab:brier_pairwise}
¥end{table}


¥section{Factor Analysis}

We can decompose the three methods into two binary factors:
¥begin{enumerate}
  ¥item ¥textbf{Web search}: on (Pro+search, Flash+search) vs.¥ off (Pro+nosearch)
  ¥item ¥textbf{Model}: Pro vs.¥ Flash (both with search)
¥end{enumerate}

¥begin{table}[h]
¥centering
¥begin{tabular}{lrr}
¥toprule
Factor & Metaculus gain & Brier gain ¥¥
¥midrule
Web search (off $¥to$ on) & $+10.9$ & $-0.0242$ ¥¥
Model (Flash $¥to$ Pro)   & $+3.4$  & $-0.0074$ ¥¥
¥bottomrule
¥end{tabular}
¥caption{Incremental gains from each factor.  Metaculus: higher is better;
  Brier: lower (more negative) is better.}
¥label{tab:factorial}
¥end{table}

The value of web search ($+10.9$ Metaculus points, equivalently $-0.024$
Brier score) is $¥sim 3¥times$ larger than the value of upgrading from
Flash to Pro ($+3.4$ points, $-0.007$ Brier).


¥section{Discussion}

¥paragraph{Web search is the dominant factor.}
Removing web search degrades Metaculus score by 11 points --- a larger
effect than the model upgrade from Flash to Pro ($+3.4$).  This suggests
that search-retrieved evidence, not parametric knowledge, is the primary
driver of forecasting accuracy on these market questions.

¥paragraph{Pro vs.¥ Flash.}
Pro outperforms Flash significantly ($p = 0.011$ Metaculus, $p = 0.035$
Brier), but the effect is modest.  Given that Flash costs $¥sim 4¥times$
less per token, Flash may offer better cost-efficiency for budget-constrained
deployments, especially when combined with multi-trial aggregation.

¥paragraph{Question difficulty dominates.}
As in the AIBQ2 analysis, question difficulty explains the vast majority
of variance ($84¥%$).  Whether a question is ``easy'' or ``hard'' matters
far more than which method is used.  The standard deviation of question
effects ($¥sigma_¥gamma = 60.2$ Metaculus points) dwarfs the method spread
($72.3 - 61.4 = 10.9$ points).

¥paragraph{Comparison with AIBQ2.}
The AIBQ2 analysis found method effects not significant ($F = 1.28$, $p > 0.05$)
because all four methods used the same search engine and model --- they
differed only in thinking level and calibration.  Here, the methods span
a wider range (with/without search, different models), yielding a highly
significant method effect ($F = 822$, $p < 0.001$).

¥paragraph{Statistical power.}
With $n = 263$ questions (vs.¥ 113 for AIBQ2), we have substantially more
power to detect effects.  The smallest significant pairwise difference
(Flash vs.¥ Pro, $¥Delta = 3.4$ Metaculus points) would not have been
detectable with 113 questions at $¥alpha = 0.05$.

¥end{document}
