\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, hyperref}
\usepackage[margin=1in]{geometry}

\title{Shrinkage Aggregation for Multi-Trial Forecasts}
\author{Kevin Murphy (with help from Claude Code)}
\date{5 April 2026}

\begin{document}
\maketitle

\section{Setup}

We run an LLM-based forecasting agent $K$ times (trials) on the same
binary question $j$, producing forecasts $p_{j,1}, \ldots, p_{j,K} \in (0,1)$.
We want to aggregate these into a single forecast $\hat{p}_j$ that performs
well under a log-scoring rule such as the Metaculus baseline score
\cite{metaculus_scoring}:
\begin{equation}
  S(p, o) = 100 \bigl(1 + \log_2 q\bigr),
  \quad q = \begin{cases} p & \text{if } o=1, \\ 1-p & \text{if } o=0, \end{cases}
\end{equation}
where $o \in \{0,1\}$ is the binary outcome and $p \in (0,1)$ is the forecast.\footnote{%
  In our implementation, $q$ is clipped to $[0.001, 0.999]$ to avoid numerical
  issues with $\log_2(0)$.}
An always-0.5 forecaster scores $S = 0$ on every question.
This scoring function penalizes overconfident wrong predictions catastrophically:
predicting $p=0.02$ when $o=1$ yields $S \approx -464$.

\section{Logit-Normal Model}

We model the trial forecasts in logit space.  Let $y_{j,k} = \mathrm{logit}(p_{j,k})$
and let $\theta_j$ denote the ``true'' logit-forecast for question $j$.

\paragraph{Likelihood.}
\begin{equation}
  y_{j,k} \mid \theta_j \;\sim\; \mathcal{N}(\theta_j,\, \sigma_j^2),
  \quad k = 1, \ldots, K,
\end{equation}
where $\sigma_j^2$ captures within-question trial-to-trial noise (stochasticity
from the LLM and web search).  We estimate it as the sample variance:
\begin{equation}
  \hat{\sigma}_j^2 = \frac{1}{K-1} \sum_{k=1}^{K} (y_{j,k} - \bar{y}_j)^2,
  \quad \bar{y}_j = \frac{1}{K} \sum_{k=1}^{K} y_{j,k}.
\end{equation}

\paragraph{Prior.}
We place a Gaussian prior centered at logit(0.5) = 0, representing an
uninformative base rate:
\begin{equation}
  \theta_j \;\sim\; \mathcal{N}(0,\, \tau^2).
\end{equation}
The hyperparameter $\tau^2$ controls the strength of shrinkage toward 0.5.

\paragraph{Posterior.}
The posterior is $\theta_j \mid y_{j,1:K} \sim \mathcal{N}(m_j, v_j)$ with
\begin{align}
  v_j &= \left(\frac{K}{\sigma_j^2} + \frac{1}{\tau^2}\right)^{-1}, \\
  m_j &= v_j \cdot \frac{K \bar{y}_j}{\sigma_j^2}
       = \alpha_j \cdot \bar{y}_j + (1 - \alpha_j) \cdot 0,
\end{align}
where the shrinkage factor is
\begin{equation}
  \alpha_j = \frac{K \tau^2}{K \tau^2 + \sigma_j^2}.
  \label{eq:alpha}
\end{equation}

\paragraph{Aggregated forecast.}
The posterior mean forecast in probability space is
\begin{equation}
  \hat{p}_j = \mathrm{sigmoid}(m_j) = \mathrm{sigmoid}(\alpha_j \cdot \bar{y}_j).
\end{equation}
When trials agree ($\sigma_j^2$ small), $\alpha_j \approx 1$ and $\hat{p}_j \approx
\mathrm{sigmoid}(\bar{y}_j)$, the plain mean in logit space.
When trials disagree ($\sigma_j^2$ large), $\alpha_j < 1$ and $\hat{p}_j$ is
pulled toward $0.5$.

\section{Practical Implementation}

Our implementation uses a heuristic that directly mirrors the posterior mean
structure of \eqref{eq:alpha}, operating in logit space:
\begin{equation}
  \hat{p}_j = \mathrm{sigmoid}(a_j \cdot \bar{y}_j),
  \quad a_j = \max\!\bigl(0.3,\; 1 - 0.7 \cdot s_j\bigr),
  \label{eq:heuristic}
\end{equation}
where $\bar{y}_j = \frac{1}{K}\sum_k \mathrm{logit}(p_{j,k})$ is the mean
logit-forecast and $s_j = \mathrm{std}(\mathrm{logit}(p_{j,1:K}))$ is the
sample standard deviation of the logit-transformed trial forecasts.

This is structurally equivalent to the Bayesian posterior mean \eqref{eq:alpha}
with $\mu = 0$:
\begin{itemize}
  \item $a_j$ plays the role of $\alpha_j = K\tau^2 / (K\tau^2 + \sigma_j^2)$.
  \item Both shrink $\bar{y}_j$ toward $0$ (i.e., $\hat{p}_j$ toward $0.5$).
  \item Both apply less shrinkage when trials agree (low $s_j$).
\end{itemize}

The key differences from the exact EB estimator:
\begin{enumerate}
  \item \textbf{Adaptive $\tau^2$}: the EB model uses a fixed $\tau^2$
    estimated globally.  The heuristic's formula $a_j = 1 - 0.7 \cdot s_j$
    implicitly adapts $\tau^2$ per question based on the local disagreement,
    providing stronger shrinkage precisely for questions where the agent is
    most uncertain.
  \item \textbf{Floor}: the $\max(0.3, \cdot)$ ensures a minimum trust in
    the data even for highly noisy questions, preventing complete collapse
    to 0.5.
  \item \textbf{Coefficients}: the constants $0.3$ (floor) and $0.7$ (scale)
    were chosen by grid search to maximize the Metaculus score on AIBQ2
    (see Section~\ref{sec:empirical}).
\end{enumerate}

\section{Empirical Comparison}
\label{sec:empirical}

On the AIBQ2 benchmark (113 questions, $K=5$ trials, Gemini 3.1 Pro):

\begin{center}
\begin{tabular}{lcc}
  \hline
  Method & Metaculus Score & Catastrophic ($< -200$) \\
  \hline
  Plain mean ($a = 1$) & 35.4 & 6 \\
  EB logit-normal ($\tau^2 = 0.10$) & 37.1 & 4 \\
  EB logit-normal ($\tau^2 = 0.15$) & 37.3 & 4 \\
  \textbf{Heuristic} \eqref{eq:heuristic} (logit-space) & \textbf{37.7} & \textbf{4} \\
  \hline
\end{tabular}
\end{center}

The heuristic slightly outperforms the best-tuned EB estimator, likely because
it adapts $\tau^2$ per question rather than using a global value.
The logit-space formulation is both more principled (matching the model) and
empirically better than the probability-space variant (37.7 vs 37.3),
since it provides appropriate nonlinear shrinkage at the extremes.

\section{Relationship to James--Stein Estimation}

The shrinkage toward 0.5 is analogous to James--Stein shrinkage
\cite{stein1956,efron1973}, which shows that shrinking individual estimates
toward a common mean reduces total MSE when estimating $\geq 3$ parameters
simultaneously.  In our setting, the ``parameters'' are the per-question
$\theta_j$ values, and we shrink each toward the uninformative prior
$\theta = 0$ (i.e., $p = 0.5$).

However, unlike classical James--Stein (which minimizes squared error), our
heuristic is implicitly optimized for log-scoring rules where the cost of
overconfident wrong predictions is catastrophic.  This makes stronger
shrinkage (smaller $\tau^2$) beneficial even when the EB MSE-optimal
estimate would suggest minimal shrinkage.

\begin{thebibliography}{9}
\bibitem{stein1956}
C.~Stein.
\newblock Inadmissibility of the usual estimator for the mean of a
  multivariate normal distribution.
\newblock In \emph{Proc.\ Third Berkeley Symposium}, 1956.

\bibitem{efron1973}
B.~Efron and C.~Morris.
\newblock Stein's estimation rule and its competitors---an empirical Bayes
  approach.
\newblock \emph{Journal of the American Statistical Association},
  68(341):117--130, 1973.

\bibitem{metaculus_scoring}
Metaculus.
\newblock Baseline score.
\newblock \url{https://www.metaculus.com/help/scores-faq/#baseline-score}, 2025.
\end{thebibliography}

\end{document}
